---
title: "Income Inequality Before and After Taxes"
about: marquee
description: "Analysis of income inequality from Tidy Tuesday"
execute:
  echo: true
  message: false
  warning: false
author: Andy Xiao
format: 
    html:
        code-fold: true
        code-summary: "Show code"


---
In this dataset we are analyzing income inequality before and after taxes. To measure inequality the Gini coefficient measures inequality on a scale from 0 to 1 with high values indicating higher inequality.

```{r}
library(tidyverse)
income_inequality_processed <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-08-05/income_inequality_processed.csv')
world_difference <- income_inequality_processed |>
    mutate(difference = gini_mi_eq - gini_dhi_eq) |>
    filter(!is.na(difference))

top5 <- world_difference|>
    group_by(Entity) |>
    summarize(mean_diff = mean(difference)) |>
    slice_max(mean_diff, n = 5)|>
    pull(Entity)

world_difference1 <- world_difference |>
    filter(Entity %in% top5)


USA_income = income_inequality_processed |>
    filter(Entity == "United States")

ggplot(world_difference1, aes(x = Year, y = difference, color = Entity))+ 
    geom_point() +
    geom_line() + 
    labs(
        title = "Top 5 countries with largest after tax difference",
        y = "post-tax Gini difference"
    )
```

In these five countries, taxes and government benefits have the greatest reduction in income inequality with most countries lowering the Gini coeffient by around 0.2 points after taxes and benefits.

```{r}
ggplot(USA_income, aes(x = Year)) +
    geom_line(aes(y = gini_dhi_eq, color = "After Tax")) + 
    geom_line(aes(y = gini_mi_eq, color = "Before Tax")) +
    labs(
        color = "Income Measure",
        y = "Income Inequality",
        title = "United States Income Inequality"

    )
```

Here is the income inequality of the United States before and after tax from 1960-2020. Here we can see a large reduction in the Gini coeffient from pre tax to after tax. This indicates that the United States tax and benefits substantially helps lower income inequality.

TidyTuesday Data set: https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-08-05/readme.md

Original source: Hasell, J. (2023). Income inequality before and after taxes: How much do countries redistribute income? Our World in Data. https://ourworldindata.org/income-inequality-before-and-after-taxes